Performing and evaluating graph-based clustering with rOpenScPCA

### 2024-11-14 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
WIP: rearranging notebook

---
title: "Performing and evaluating graph-based clustering with rOpenScPCA"
date: "`r Sys.Date()`"
author: Data Lab
params:
  seed: 2024
output: 
  html_notebook: 
    toc: yes
    toc_float: yes
    df_print: paged
---

@@

- Perform clustering on an SCE or Seurat object
- Evaluate clustering results using several metrics including:
  - Silhouette width
  - Neighborhood purity
  - Cluster stability, as measured with the Adjusted Rand index
- Save clustering results to an SCE or Seurat object

@@

Because clustering involves random samples, it is important to set the random seed at the top of your analysis script or notebook to ensure reproducibilty.

@@

### Input and output files

@@

To begin, we'll read in the `SingleCellExperiment` (SCE) object and convert it to a Seurat object.
This will allow us to demonstrate `rOpenScPCA` usage with both SCE and Seurat objects.

@@

# Convert to a Seurat object for demonstration
seurat_obj <- as.Seurat(
  sce,
  counts = "counts",
  data = "logcounts"
)
```

@@

By default, this function will calculate clusters using the following parameters:

@@

These parameters can be customized, as follows:

@@

* `cell_id`: Unique cell identifiers, e.g. cell barcodes
* `cluster`: A factor column with the cluster identities
* There will be one column for each clustering parameter used

### Clustering on an SCE object

```{r cluster sce}
# Calculate clusters with default parameters
cluster_results <- calculate_clusters(sce)

# Print resulting table
cluster_results
```

The following chunk shows an example of performing clustering with several non-default parameters:

```{r cluster sce nondefault}
# Calculate clusters with non-default parameters
cluster_results <- calculate_clusters(
  sce,
  algorithm = "leiden",
  nn = 15,
  objective_function = "modularity",
  seed = params$seed
)

# Print resulting table
cluster_results
```

```{r cluster seurat}
cluster_results <- calculate_clusters(
  seurat_obj,
  # Since we converted this from an SCE object, its PC name is `PCA`
  pc_name = "PCA"
)

cluster_results
```

@@

This section demonstrates how to use several functions for evaluating cluster quality and reliability. 
All examples will use the SCE object for demonstration, but a Seurat object can be provided instead.

@@

* An SCE or Seurat object that contains principal components
  * If the PCA matrix name is not the default for the object type (`PCA` for SCE and `pca` for Seurat), you can use `pc_name` to specify its name.
* A data frame with, at least, columns representing unique cell ids and cluster assignments.
  * By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters()`
  * You can override these defaults using the arguments `cell_id_col` and `cluster_col`

@@

We'll use the function `rOpenScPCA::calculate_silhouette()` to calculate the silhouette width, and then we'll visualize the results.

@@

```{r silhouette}
silhouette_results <- calculate_silhouette(
  sce,
  cluster_results
)

silhouette_results
```

```{r plot silhouette density}
# Calculate mean silhouette width
mean_silhouette_width <- mean(silhouette_results$silhouette_width)

# Visualize values as a density plot, showing the mean value as a vertical line
ggplot(silhouette_results) +
  aes(x = silhouette_width) +
  geom_density(fill = "lightblue") +
  geom_vline(xintercept = mean_silhouette_width, color = "red", linewidth = 1)
```

Looking at the full distribution of silhouette width values, we see most are positive which is encouraging, but the values are mostly below 0.5, and the mean is less than 0.25.
Silhouette width therefore suggests that, while these clusters do have some structure, they are not very tight.

We can also explore the silhouette width results on a per-cluster basis:

```{r plot silhouette violin}
ggplot(silhouette_results) +
  aes(x = cluster, y = silhouette_width) +
  geom_violin() +
  geom_jitter(size = 0.75, alpha = 0.25) +
  # show the mean +/- SE in red
  stat_summary(color = "red")
```

This visualization shows us that cluster 5 in particular has the lowest silhouette width values, and clusters 11 and 12 have the highest.

@@

We'll use the function `rOpenScPCA::calculate_purity()` to calculate the neighborhood purity, and then visualize the results.

@@

```{r purity}
purity_results <- calculate_purity(
  sce,
  cluster_results
)

purity_results
```

```{r plot purity density}
mean_purity <- mean(purity_results$purity)

# Visualize values as a density plot, showing the mean value as a vertical line
ggplot(purity_results) +
  aes(x = purity) +
  geom_density(fill = "lightblue") +
  geom_vline(xintercept = mean_purity, color = "red", linewidth = 1)
```
This metric shows a relatively high average neighborhood purity, suggesting that neighboring cells tend to belong to the same cluster.

We can again visualize this on a per-cluster basis:

```{r plot purity violin}
ggplot(purity_results) +
  aes(x = cluster, y = purity) +
  geom_violin() +
  geom_jitter(size = 0.75, alpha = 0.25) +
  # show the mean in value
  stat_summary(color = "red")
```

In this view, we see that clusters 1, 3, and 6 all have average values below 0.75, but the rest are higher, in particular clusters 11, 12, and 13.
It's worth noting that both silhouette width and neighborhood purity found that clusters 11 and 12 were among the tightest clusters.

@@

We'll use the function `rOpenScPCA::calculate_stability()` to calculate the stability of clusters using a bootstrapping approach, and then visualize the results.
By default, this function performs 20 bootstrap replicates, but this can be customized using the argument `replicates`.

@@

Since this function uses random sampling when performing bootstrapping, it's also important to pass in a random seed for reproducibility with the argument `seed`.

```{r stability, warning=FALSE}
stability_results <- calculate_stability(
  sce,
  cluster_results,
  seed = params$seed
)

stability_results
```

We can visualize these results, for example, with a density plot with a rug to emphasize individual values:

```{r plot stability}
ggplot(stability_results) +
  aes(x = ari) +
  geom_density(fill = "lightblue") +
  geom_rug(color = "red", linewidth = 1)
```

All bootstrap replicates produced values around 0.7, which is relatively high and close to 1, which suggests that the these clusters are stable and robust to starting conditions.

@@

```{r stability custom parameters, warning=FALSE}
# Calculate clusters with non-default parameters
cluster_df_leiden <- calculate_clusters(
  sce,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15,
  seed = params$seed
)

# Now, pass in the same arguments customizing parameters here
stability_results_leiden <- calculate_stability(
  sce,
  cluster_df_leiden,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15,
  seed = params$seed
)

@@

@@

Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your scRNA-seq object.
The subsequent examples will show you, as an example, how to save the `cluster_results` data frame to a given object.

@@

To avoid ambiguity between the existing and new clustering results, you may want to rename incoming columns.
We'll therefore begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results` data frame before adding them to a given object.

```{r rename columns}
# First, rename columns in `cluster_results` to avoid ambiguity
cluster_results_renamed <- cluster_results |>
  # add the prefix openscpca_ to all columns
  dplyr::rename_with(~ paste0("ropenscpca_", .x))

cluster_results_renamed
```

@@

# Third, add columns as desired using the barcodes and cell_id columns to join
sce_coldata <- sce_coldata |>
  dplyr::left_join(
    cluster_results_renamed,
    by = c("barcodes" = "ropenscpca_cell_id")
  )

@@

```{r add to seurat}
# Add the cluster result data frame to the Seurat object metadata
seurat_obj <- AddMetaData(seurat_obj, cluster_results_renamed)

>>>>

---
title: "Performing and evaluating graph-based clustering with rOpenScPCA"
date: "`r Sys.Date()`"
author: "Data Lab"
params:
  seed: 2024
output: 
  html_notebook: 
    toc: yes
    toc_float: yes
    df_print: paged
---

@@

- Perform clustering
- Evaluate clustering results using several metrics, including:
  - Silhouette width
  - Neighborhood purity
  - Cluster stability, as measured with the Adjusted Rand Index
- Save clustering results

While this notebook does demonstrate how to use individual functions that evaluate a single clustering result, a full clustering evaluation would entail comparison these metrics across different clusterings from different parameterizations.

@@

Because clustering involves random sampling, it is important to set the random seed at the top of your analysis script or notebook to ensure reproducibilty.

@@

### Input file

@@

To begin, we'll read in the `SingleCellExperiment` (SCE) object.
We'll also extract its PCA matrix, which we'll use to demonstrate how to calculate and evaluate clusters.

@@

# Extract the PCA matrix
pca_matrix <- reducedDim(sce, "PCA")
```

@@

This function takes a PCA matrix with rownames representing unique cell ids (e.g., barcodes) as its primary argument.
By default it will calculate clusters using the following parameters:

@@

* `cell_id`: Unique cell identifiers, obtained from the PCA matrix's row names
* `cluster`: A factor column with the cluster identities
* There will be one column for each clustering parameter used

### Clustering with default parameters

```{r cluster sce}
# Calculate clusters with default parameters
cluster_results_df <- calculate_clusters(pca_matrix)

# Print resulting table
cluster_results_df
```

### Clustering with non-default parameters

Parameters used for clustering can be customized with these arguments:

@@

For example:

```{r cluster sce nondefault}
# Calculate clusters with non-default parameters
cluster_results_df <- calculate_clusters(
  pca_matrix,
  algorithm = "leiden",
  nn = 15,
  objective_function = "modularity"
)

# Print resulting table
cluster_results_df
```

@@

This section demonstrates how to use several functions for evaluating cluster quality and reliability. 
It's important to note that a full evaluation of clustering results would compare these metrics across a set of clustering results, with the aim of identifying an optimal parameterization.

@@

* A PCA matrix with rownames representing unique cell ids (e.g., barcodes)
* A data frame with, at least, columns representing unique cell ids and cluster assignments
  * By default, these columns should be named `cell_id` and `cluster`, respectively, matching the output of `rOpenScPCA::calculate_clusters()`
  * You can override these defaults using the arguments `cell_id_col` and `cluster_col`

@@

We'll use the function `rOpenScPCA::calculate_silhouette()` to calculate the silhouette width.

@@

```{r silhouette}
# calculate the silhouette width for each cell
silhouette_results <- calculate_silhouette(
  pca_matrix,
  cluster_results_df
)

# print result
silhouette_results
```

@@

We'll use the function `rOpenScPCA::calculate_purity()` to calculate the neighborhood purity.

@@

```{r purity}
# calculate the neighborhood purity for each cell
purity_results <- calculate_purity(
  pca_matrix,
  cluster_results_df
)

# print result
purity_results
```

@@

We'll use the function `rOpenScPCA::calculate_stability()` to calculate the cluster stability.
By default, this function performs 20 bootstrap replicates, but this can be customized using the argument `replicates`.

@@

```{r stability, warning=FALSE}
# calculate the stability of clusters
stability_results <- calculate_stability(
  sce,
  cluster_results_df
)

# print the result
stability_results
```

@@

```{r stability custom parameters}
# Calculate clusters with non-default parameters
cluster_df_leiden <- calculate_clusters(
  sce,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

# Now, pass in the same arguments customizing parameters here
stability_results_leiden <- calculate_stability(
  sce,
  cluster_df_leiden,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

@@

## Working with objects directly

The above examples present how to work with a pca matrix, but you can also pass in an SCE or Seurat object with PCs to use instead.
When doing so, this package will assume a pc_name, and you will need to override this as needed.

```{r}
# Convert to a Seurat object for demonstration
seurat_obj <- as.Seurat(
  sce,
  counts = "counts",
  data = "logcounts"
)
```

```{r cluster seurat}
cluster_results_df <- calculate_clusters(
  seurat_obj,
  # Since we converted this from an SCE object, its PC name is `PCA`
  pc_name = "PCA"
)

cluster_results_df
```

@@

Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your scRNA-seq object.
The subsequent examples will show you, as an example, how to save the `cluster_results_df` data frame to a given object.

@@

To avoid ambiguity between the existing and new clustering results, you may want to rename incoming columns.
We'll therefore begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results_df` data frame before adding them to a given object.

```{r rename columns}
# First, rename columns in `cluster_results_df` to avoid ambiguity
cluster_results_renamed_df <- cluster_results_df |>
  # add the prefix openscpca_ to all columns
  dplyr::rename_with(~ paste0("ropenscpca_", .x))

cluster_results_renamed_df
```

@@

# Third, add columns as desired using the barcodes and cell_id columns to join
sce_coldata <- sce_coldata |>
  dplyr::left_join(
    cluster_results_renamed_df,
    by = c("barcodes" = "ropenscpca_cell_id")
  )

@@

```{r add to seurat}
# Add the cluster result data frame to the Seurat object metadata
seurat_obj <- AddMetaData(seurat_obj, cluster_results_renamed_df)
